Quantum H theorem, variational entropy, and relaxation processes

### 2021-01-10 tex/classical.tex
Rewording: start of sec. 3.2.1
Los los cambios que hizo el Dr. Carrillo hasta a página 10 fueron agregados

Classical statistical mechanics is theoretically built upon the Boltzmann transport
equation and the hypothesis of molecular chaos. The fundamental connection
between the microscopic nature of matter and the phenomenological laws of thermodynamics
is the $H$-theorem. In this article we propose an alternative approach, which is 
developed by means of the variational method, applied to a functional that may include
systems out-of-equilibrium (or non-homogeneous). Our proposal is limited to systems
whose initial state is physically close to the equilibrium, in such a manner that the
local equilibrium hypothesis is satisfied.

We begin this section by quite briefly stating the Boltzmann transport equation, followed
by an also quite brief review of the $H$-theorem demonstration, as performed in
standard textbooks.

@@

From the Boltzmann kinetic theory of gases, the probability for a particle to be in a given
state of position and velocity at a time $t$,
$f(\vec r,\vec v,t)$ (\textit{aka} the distribution function), must satify
the transport equation (Eq.~(\ref{eq:transport}), below).
This equation is traditionally obtained by exploiting the volume-invariance of the $\mu$-space
(\textit{i.e.} the space formed by the direct product of position and velocity)
and considering that elastic collisions constitute the predominant mechanism
for exchanging energy among particles \cite{bib:huang}.
%
\begin{equation}\label{eq:transport}
	%\left(\frac{\partial f_1}{\partial t}\right)_{\mathrm{coll}}=
  \left(
    \frac{\partial}{\partial t}+\vec{v}_1 \cdot\nabla_{\vec r}
    +\frac{\vec{F}}{m}\cdot\nabla_{\vec v_1}
  \right)f_1=\int\mathrm{d}\Omega\int
    \mathrm{d}^{3}v_2\sigma(\Omega)|\vec{v}_1-\vec{v}_2|(\tilde f_2\tilde f_1-f_2f_1).
\end{equation}
%
In Eq. (\ref{eq:transport}), $\Omega$ is the solid angle, $\sigma$ is the
scattering cross section, $\vec F$ the
external force applied to the system, and $f_1$ and $f_2$ ($\tilde f_1$ and $\tilde f_2$) are the distribution
functions of particles 1 and 2, respectively, before (after) the collision.

Particle dynamics and external forces are described
by the left-hand side of Eq.~(\ref{eq:transport}), and the right-hand side
is derived considering binary collisions between particles and
the \textit{molecular chaos hypothesis}, \textit{i.e.} it is assumed that
the coordinates and velocities of the particles are not time-correlated.

@@

The evolution of a dilute gas towards thermodynamic equilibrium is frequently
addressed by first defining the functional \cite{bib:tolman,bib:huang}
%
\begin{equation}\label{eq:hbfunctional}
   H_{B}=\int f(\vec{v},t) \ln f(\vec{v},t) \mathrm{d}^{3}v.
\end{equation} 
%
Here, $f(\vec{v},t)$ is a homogeneous distribution function, \textit{i.e.}
$f$ is assumed to be independent of $\vec r$.
The functional $H_B$,
originally introduced by Boltzmann in 1872,
describes a diluted gas, at temperature $T$, with a total energy $E$, a total
number of free classical particles $N$, and occupying a volume $V$.
(We deviate from the custommary notation in order to
clearly distinguish the Boltzmann functional $H$ from our proposals, and
$f_B$ to denote the Maxwell-Boltzmann distribution function.)

Under this scheme, the time-evolution of the gas towards equilibrium is obtained through the
$H$-theorem. This theorem establishes that if $f(\vec{v},t)$ satisfies the Boltzmann
transport equation and the molecular chaos hypothesis is valid, then
$dH_B/dt\leq0$ in general, and if $dH_B/dt=0$, then the system is in the equilibrium state.
The $H$-theorem is proven by taking the time derivative
of Eq. (\ref{eq:hbfunctional}) and using the Boltzmann transport equation
(\ref{eq:transport}) \cite{bib:huang}. The $H$-theorem fixes the direction of time-evolution
of systems in nature, which is consistent with the second law of thermodynamics; in fact,
$H_B$ can be associated with an entropy density.

On the other hand, considering a diluted gas in equilibrium (\textit{i.e.}
$(\partial f/\partial t)=0$),
with no applied external forces (\textit{i.e.} $f$ is independent of $\vec r$), 
it can be proven that the equilibrium distribution function obtained from
Eq.~(\ref{eq:transport}) is precisely the Maxwell-Boltzmann
distribution function. The proof of the above commences by finding the sufficient
condition for $f$ to render a null r.h.s. of Eq. (\ref{eq:transport}). Such an
$f$, which we denote here as $f_0$, must satisfy:
%
\begin{equation}\label{eq:neccondf0}
    f_0(\vec{v}'_2)f_0(\vec{v}'_1)- f_0(\vec{v}_2)f_0(\vec{v}_1)=0.
\end{equation}
%
Subsequently, the Maxwell-Boltzmann distrubution function can be obtained by
taking the logarithm to Eq.~(\ref{eq:neccondf0}), and exploiting mechanical
conserved quantities (see \cite[ch. 4.2]{bib:huang}).

Before introducing our proposal, we bring the reader's attention to the following. In defining
Eq.~(\ref{eq:hbfunctional}), it is assumed that the distribution function does not
depend on the coordinates, which implies that the initial state of the system may not in
equilibrium, but the difference between the equilibrium and non-equilibrium is such
that it does not cause inhomogeneities to the system.

%---------------------------
\subsection{Alternative classical $H$-functional}

As we saw in the previous section,
the validity of the $H$-theorem relies significantly on assuming
that the distribution function is homogeneous and that the hypothesis of
molecular chaos is surmised. However, in order to consider systems whose
distribution functions are not necessarily homogeneous, which
allows for studying systems out of equilibrium,
we introduce a modified $H$-functional as follows. Hereafter, we will
use the term homogeneous distribution function to refer to a distribution
function thad does not depend on the coordinates. In addition,
in this section we will use primed functions
and quantities to differentiate the classical case (this Section) \textit{vs} the quantum
case (below, Section \ref{sec:quantum}).

Our proposed functional, denoted as $\mathcal{H}'$, describes a diluted classical
gas occupying a volume, $V$,
divided into $K$ cells, which, without loss of generality,
have local identical volumes (\textit{i.e.} $\delta V_M = V/K,\ M=1,\,\dots\,,K$).
Each cell has its own local
$H$-functional, $\mathcal{H}'_M$, and local homogeneous distribution function,
$f'_{M}(\vec{v},t)$; local number of particles, $\mathcal{N}'_M$;
local temperature, $T'_M$; and a local energy, $\mathcal{E}'_M$. As a whole, the system has a
global energy $E$, and a global number of free
classical particles $N$, and it is considered to be perfectly isolated (\textit{i.e.} the
global system is a microcanonical ensemble). We also assume that the number of particles
in each cell is sufficiently large, so as to obtain accurate averages.
The functional takes the form
%
\begin{equation}\label{eq:CH2}
   \mathcal{H}'(t)=\sum_{M=1}^{K}\int_{\delta V_M} f'_M(\vec{v},t) \ln f'_M(\vec{v},t)\mathrm{d}^3v.
\end{equation}
%

The distribution functions, $\{f'_M(\vec{v},t)\}$, depend implicitly on the position of the cells,
relative to the global system, and on the velocity
$\vec{v}$ and time $t$. Each $f'_M$ can be formally extended to the complete coordinate space
by defining each $f'_M$ to be zero outside the $M$-th cell, in such a manner that
the distribution function of the complete system is a piece-wise sum
of $\{f'_M(\vec v,t)\}$:
%
\begin{equation}
   f'(\vec r,\vec v,t)=\sum_{M=1}^Kf'(\vec r_M,\vec v,t)=\sum_{M=1}^Kf'_M(\vec v,t).
\end{equation}
%
Here $\vec r_M$ is the center of the cell of index $M$, and $f'_M(\vec v,t)\neq f'_N(\vec v,t)$
for $M\neq N$. This extended definition
allows us to omit the symbol $\delta V_M$ on all integrals performed over the cell volume.
In terms of $f'_M(\vec{v},t)$ and the energy, $\epsilon(\vec{v})$, we have:
%
\begin{subequations}\label{eq:cellrestrictions}
\begin{eqnarray}
    \mathcal{H}_M' & = &  \int f'_M(\vec{v},t) \ln f'_{M}(\vec{v},t)
      \mathrm{d}^{3}v \label{eq:Hcell},\\
    \mathcal{N}_M' & = & \int f'_{M}(\vec{v} ,t) \mathrm{d}^{3}v,\label{eq:Ncell}\\
    \mathcal{E}_M' & = & \int f'_{M}(\vec{v},t)\epsilon(\vec{v}) \mathrm{d}^{3}v\label{eq:Ecell}.
\end{eqnarray}
\end{subequations}
%
Notice that assuming homogeneous distribution functions per cell means that we are surmising the 
local equilibrium hypothesis. In addition, the set $\{f'_{M}(\vec{v},t)\}$ must satisfy the following
microcanonical restrictions:
%
\begin{subequations}\label{eq:micro}
\begin{equation}\label{eq:micron}
    \sum_{M=1}^{K}\int f'_M(\vec{v},t)\mathrm{d}^3v =N
\end{equation}
and
\begin{equation}\label{eq:microe}
    \sum_{M=1}^{K}\int f'_M(\vec{v},t)\epsilon(\vec{v})\mathrm{d}^3v=E.
\end{equation}
\end{subequations}
%

We now use the variational method in order to maximize $\mathcal{H}'$, applying the restrictions
given in Eqs.~(\ref{eq:cellrestrictions}) and (\ref{eq:micro}) through Lagrange
multipliers ($\{\alpha_M\}$ and $\{\beta_M\}$). This yields
%
\begin{eqnarray}
    \frac{\delta \mathcal{H}'}{\delta f'_J(\vec{v}')} & = & \sum_{M=1}^{K}\int
      \frac{\delta}{\delta f'_J(\vec{v}')}\left[
        f'_M(\vec{v})\ln f'_M(\vec{v})
        \right]
       \mathrm{d}^3v -\sum_{M=1}^{K}\alpha_M\int
       	\frac{\delta f'_M(\vec{v})}{\delta f'_J(\vec{v}')}
      \mathrm{d}^3v\nonumber\\
    & & -\sum_{M=1}^{K}\beta_M\int\epsilon(\vec{v})
    	\frac{\delta f'_M(\vec{v})}{\delta f'_J(\vec{v}')}
      \mathrm{d}^3v \nonumber\\
    & = & \ln f'_J(\vec{v}')+1-\alpha_J-\beta_J \epsilon(\vec{v}')=0.
\end{eqnarray}
%
Solving the last line for $f'_J(\vec{v}')$ renders
%
\begin{comment}
\begin{eqnarray}
    \ln f'_J(\vec{v}') & = & \alpha_J+\beta_J \epsilon(\vec{v}')
       -1\quad\Rightarrow\quad f_J(\vec{v}')\\
    & = & e^{\alpha_J +\beta_J \epsilon(\vec{v}')-1}\nonumber\\
    & = & Ce^{\alpha_J+\beta_J \epsilon(\vec{v}') } \label{eq:relacion1},
\end{eqnarray}
\end{comment}
\begin{equation}\label{eq:relacion1}
	f'_J(\vec{v}') = C\exp\left({\alpha_J+\beta_J \epsilon(\vec{v}') }\right),
\end{equation}
%
where $C$ is a constant. We notice that using the variational procedure on
$\mathcal{H}'$ predicts that the
distribution function of each cell has the form of the Maxwell-Boltzmann
distribution function, which is consistent with the local equilibrium assumption.

@@

If the complete system is in equilibrium without external forces applied to
the gas, from classical thermodynamics of
systems in equilibrium, we ascertain that the local number
of particles and the local
energy do not depend on the cell number, on a statistical sense, this is
%
\begin{subequations}\label{eq:eqNandE}
\begin{equation}
   \mathcal{N}'_M=\mathcal{N}'\equiv \bar{\mathcal{N}}'
\end{equation}
%
and
%
\begin{equation}
   \quad\mathcal{E}'_M=\mathcal{E}'\equiv\bar{\mathcal{E}}'.
\end{equation}
\end{subequations}
%
In Eqs. (\ref{eq:eqNandE}) the bar implies averaged properties over the complete system.
Moreover, the distribution function is homogeneous, hence $f'_M$
does not depend on the cell number $M$ (\textit{i.e.} $f'_M(\vec v,t)=f'(\vec v,t)$,
$\forall M$). Several properties follow directly from this, \textit{e.g.}
from Eqs.~(\ref{eq:eqNandE}) $E=\sum_M\mathcal{E}'_M=K\bar{\mathcal{E}}'$, $N=\sum_M\mathcal{N}'_M=K\bar{\mathcal{N}}'$ (here we have used Eqs. (\ref{eq:Ecell})
and (\ref{eq:Ncell})),
and Eq.~(\ref{eq:CH2}), in terms of Eq.~(\ref{eq:hbfunctional}), reduces to
%
\begin{equation}
    \mathcal{H}'(t)=\int \sum_{M=1}^{K} [f'(\vec{v},t)\ln f'(\vec{v},t)] \mathrm{d}^{3}v
         = K\int  f'(\vec{v},t) \ln f'(\vec{v},t) \mathrm{d}^{3}v= K H_{B}(t).
\end{equation}
%
In order to identify $\mathcal{H}'$ with the entropy, it remains to show that $\mathcal{H}'(t)$ is extensive,
with respect to $Kf'_(\vec v,t)$.
This is shown by analyzing the following expression
%
\begin{eqnarray}\label{eq:sum-h}
    \int \big[Kf'(\vec{v},t)\big] \ln \big[Kf'(\vec{v},t)\big] \mathrm{d}^{3}v & = & 
        \int \big[(K\ln K)f'(\vec{v},t) + Kf'(\vec{v},t) \ln f'(\vec{v},t)\big]\mathrm{d}^{3}v\nonumber\\
        & = & K\int\Big\{ f'(\vec{v},t)\big[\ln K + \ln f'(\vec{v},t)\big] \Big\}\mathrm{d}^{3}v.
\end{eqnarray}
%
We observe that if the number of particles in the $\mu$-space, $f'(\vec{v},t)$, is much greater than
the number of cells $K$, then the first term of Eq.~(\ref{eq:sum-h}) is negligible, and consequently:
%
\begin{equation}\label{aditive-property}
    \int \big[Kf'(\vec{v},t)\big] \ln \big[Kf'(\vec{v},t)\big] \mathrm{d}^{3}v \simeq
    K\int f'(\vec{v},t) \ln f'(\vec{v},t) \mathrm{d}^{3}v,
\end{equation}
%
\textit{i.e.} $\mathcal{H}'_M$ is extensive, and the sum $\sum_M\mathcal{H}'_M$ is the $H$-functional of the complete
system, which reduces to the Boltzmann $H$-functional. Hence, $\mathcal{H}'$ can also be identified with the
system's entropy.

Furthermore, in equilibrium the Lagrange multipliers are position- and time-independent,
thus $f'_M(\vec{v})$
reduces to
%
\begin{equation}\label{eq:fbardef}
    f'_M(\vec{v})=C\exp\left(\alpha+\beta \epsilon(\vec{v})\right)
    \equiv \bar{f}'(\vec{v}),\quad M=1,\,\dots\,,K.
\end{equation} 
%
The constant $C$ can be omitted, which is shown by defining the following $\mathcal{H}''$
functional:
%
\begin{equation}
   \mathcal{H}''(t)=\sum_{M=1}^{K}\int \left[f'_M(\vec{v},t)
    \ln f'_M(\vec{v},t)-f'_M(\vec{v},t)\right]\mathrm{d}^3v  \label{CH3},
\end{equation}
%
however $\sum_{M=1}^{K} \int f'_M(\vec{v},t)\mathrm{d}^3v =N$.
Since $N$ is a constant and we are mainly intersted in the time-derivative of $\mathcal{H}''$,
$C$ can be conveniently omitted. In other words, $\mathcal{H}'$ renders the Maxwell-Boltzmann
distribution function for systems in equilibrium.

\subsubsection{Alternative proof of the classical $H$-theorem}

>>>>

Boltzmann Kinetic theory of gases is one of the theoretical basese of our present 
knowledge of out equilibrium systems. This represent a fundamental connection
between the microscopic nature of matter and the phenomenological macroscopic  laws of classical thermodynamics.
The stochasticity introduced in the, in other way deterministic, kinetics of the
particles of the gas, via the molecular chaos hypothesis, allows the demostration 
of the celebreted Boltzmann $H$ theorem. In this article we propose an alternative approach, which is 
developed by means of a variational method, applied to a functional.
Let us start this section by briefly remaind of the important elements of the 
standard derivation and demostration of the Boltzmann transport equation and the $H$ theorem as they are presented in classical textbooks \cite{bib:huang}.

@@

Boltzmann kinetic theory of gases starts by defining the distribution function 
$f(\vec{r},\vec{v},t)$ as the average number of molecules that at time $t$, has the 
values of position and velocity in the element of volumen in the $\mu$ space.
Assuming a deterministic Newtonian description of the movement of the molecules 
and the invariance of the volume measure of the $\mu$ space, one arrives to the Boltzmann rate equation.
%
\begin{equation}
	f(\vec{r}+\vec{v}\delta t, \vec{v}+\vec{F} \delta t, t+\delta t)=f(\vec{r},\vec{v},t)+\left( \frac{\partial f}{\partial t} \right)_{coll} \delta t.
\end{equation}
%
The collision term $\left( \frac{\partial f}{\partial t}\right)$ of this equation 
describes the fuxes in and out from the volume element due to the collisions.
The integro-differential Boltzmann transpor equation can be cast into.
%
\begin{equation}\label{eq:transport}
	%\left(\frac{\partial f_1}{\partial t}\right)_{\mathrm{coll}}=
  \left(
    \frac{\partial}{\partial t}+\vec{v}_1 \cdot\nabla_{\vec r}
    +\frac{\vec{F}}{m}\cdot\nabla_{\vec v_1}
  \right)f_1=\int\mathrm{d}\Omega\int
    \mathrm{d}^{3}v_2\sigma(\Omega)|\vec{v}_1-\vec{v}_2|(\tilde f_2\tilde f_1-f_2f_1).
\end{equation}
%
In Eq. (\ref{eq:transport}), $\Omega$ is the solid angle, $\sigma$ is the
scattering cross section, $\vec F$ the
external force applied to the system, and $f_1$ and $f_2$ ($\tilde f_1$ and $\tilde f_2$) are the distribution
functions of particles 1 and 2, respectively, before (after) the collision.

Particle dynamics and the effects of external forces are described
by the left-hand side of Eq.~(\ref{eq:transport}), and the right-hand side
is derived considering binary collisions between particles and accepting
the \textit{molecular chaos hypothesis}, \textit{i.e.} it is assumed that
the positions and velocities of the particles are not time-correlated.

@@

The evolution of a dilute gas towards thermodynamic equilibrium is frequently
addressed by first defining the functional \cite{bib:tolman,bib:huang}
%
\begin{equation}\label{eq:hbfunctional}
   H_{B}=\int f(\vec{v},t) \ln f(\vec{v},t) \mathrm{d}^{3}v.
\end{equation} 
%
Notice that $f(\vec{v},t)$ is a spacially homogeneous distribution function.
The functional $H_B$,
originally introduced by Boltzmann in 1872,
describes a diluted gas, at temperature $T$, with a total energy $E$, a total
number of free classical particles $N$, and occupying a volume $V$.
In order to clearly distinguish the Boltzmann functional $H_B$, we denote here
the Maxwell-Boltzmann distribution function as $f_B$.

The physically correct espontaneous time evolution of an out of equilibrium dilute gas 
is corroborated by the $H$ theorem.
This theorem establishes that if the homogeneous function $f(\vec{v},t)$ satisfies the Boltzmann
transport equation and consequently, if the molecular chaos hypothesis is valid, then
the system evolves in such a way that $dH_B/dt\leq0$, and if when $dH_B/dt=0$, 
the system is in the equilibrium state. $H$-theorem is easily proven by 
using the Boltzmann transport equation (\ref{eq:transport}) \cite{bib:huang}. 
$H$-theorem assures that our microscopic approach to describe the spontaneous 
time evolution of the system will be consistent with the phenomenological observations
established in the 2nd-law of Classical; in fact,
$H_B$ can be associated with an entropy density.

On the other hand, considering a diluted gas in equilibrium, \textit{i.e.}
$(\partial f/\partial t)=0$,
with no applied external forces $f$ is independent of $\vec r$, 
it can be directly proven that the equilibrium distribution function obtained from
Eq.~(\ref{eq:transport}) is precisely the Maxwell-Boltzmann
distribution function. The proof of the above commences by finding the sufficient
condition for $f$ to render a null r.h.s. of Eq. (\ref{eq:transport}). Such an
$f$, which we denote here as $f_0$, must satisfy:
%
\begin{equation}\label{eq:neccondf0}
    f_0(\vec{v}'_2)f_0(\vec{v}'_1)- f_0(\vec{v}_2)f_0(\vec{v}_1)=0.
\end{equation}
%
Subsequently, the Maxwell-Boltzmann distrubution function can be obtained by
taking the logarithm to Eq.~(\ref{eq:neccondf0}), and exploiting mechanical
conserved quantities (see \cite[ch. 4.2]{bib:huang}).

Before introducing our proposal, we call the reader's attention to the following. In defining
$H_B$ Eq.~(\ref{eq:hbfunctional}), it is assumed that the distribution function 
$f_B$ is spattially homogeneous. Clearly this assumption simplifies the demostration 
of the $H$ theorem, nevertheless, it introduces as well a limitate-conception
of an out of equilibrium condition in the gas. Given the relatively simple nature of a dilute 
gas, one of the main characteristics of an out of equilibrium condition would be the existence of
inhomogeneities in the system.

%---------------------------
\subsection{Non-homogeneous classical $H$-functional}

As we saw in the previous section,
the validity of the $H$-theorem relies significantly on assuming
that the distribution function is homogeneous and that the hypothesis of
molecular chaos is fullfiled. In order to extend the procedure for systems whose
distribution functions are not necessarily homogeneous, would possibilities allows 
for studying systems been in a more general out of equilibrium condition, we introduce a modified $H$-functional 
as follows. For the sake of clearity and simplicity we will use primed functions
and quantities to denotate the classical case and differentiate it from the quantum
case.

Our proposed functional, denoted as $\mathcal{H}'$, describes a dilute classical
gas occupying a volume $V$. For our theoretical treatment we will consider this volume
divided into $K$ cells, which, without loss of generality,
have identical volumes $\delta V_M = V/K,\ M=1,\,\dots\,,K$.
Locally the $\mathcal{H}'$-functional is denoted by $\mathcal{H}'_M$, and 
the local homogeneous distribution function, $f'_{M}(\vec{v},t)$; local 
number of particles, $\mathcal{N}'_M$; local temperature, $T'_M$; and a local 
energy, $\mathcal{E}'_M$. As a whole, the system has a
energy $E$, and a global number of free classical particles $N$. Let us assume that the system is 
to be isolated. We also assume that the number of particles
in each cell is sufficiently large, so as to obtain accurate averages.
We start our analysis by proposing the following inhomogeneous functional
%
\begin{equation}\label{eq:CH2}
   \mathcal{H}'(t)=\sum_{M=1}^{K}\int_{\delta V_M} f'_M(\vec{v},t) \ln f'_M(\vec{v},t)\mathrm{d}^3v.
\end{equation}
%

The distribution functions, $\{f'_M(\vec{v},t)\}$, depend implicitly on the position of the cells,
relative to the global system, and on the velocity
$\vec{v}$ and time $t$. Notice that each $f'_M$ can be formally extended to the complete coordinate space
by defining each $f'_M$ to be zero outside the $M$-th cell, in such a manner that
the distribution function of the complete system is a piece-wise sum
of $\{f'_M(\vec v,t)\}$:
%
\begin{equation}
   f'(\vec r,\vec v,t)=\sum_{M=1}^Kf'(\vec r_M,\vec v,t)=\sum_{M=1}^Kf'_M(\vec v,t).
\end{equation}
%
Here $\vec r_M$ is the center of the cell of index $M$, and $f'_M(\vec v,t)\neq f'_N(\vec v,t)$
for $M\neq N$. This extended definition
allows us to omit the symbol $\delta V_M$ on all integrals performed over the cell volume.
In terms of $f'_M(\vec{v},t)$ and a local variable of energy, $\epsilon(\vec{v})$, we have:
%
\begin{subequations}\label{eq:cellrestrictions}
\begin{eqnarray}
    \mathcal{H}_M' & = &  \int f'_M(\vec{v},t) \ln f'_{M}(\vec{v},t)
      \mathrm{d}^{3}v \label{eq:Hcell},\\
    \mathcal{N}_M' & = & \int f'_{M}(\vec{v} ,t) \mathrm{d}^{3}v,\label{eq:Ncell}\\
    \mathcal{E}_M' & = & \int f'_{M}(\vec{v},t)\epsilon(\vec{v}) \mathrm{d}^{3}v\label{eq:Ecell}.
\end{eqnarray}
\end{subequations}
%
Notice that assuming homogeneous distribution functions per cell means that we are accepting valid the 
local equilibrium hypothesis. In addition, the set $\{f'_{M}(\vec{v},t)\}$ must satisfy the following restrictions:
%
\begin{subequations}\label{eq:micro}
\begin{equation}\label{eq:micron}
    \sum_{M=1}^{K}\int f'_M(\vec{v},t)\mathrm{d}^3v =N
\end{equation}
and
\begin{equation}\label{eq:microe}
    \sum_{M=1}^{K}\int f'_M(\vec{v},t)\epsilon(\vec{v})\mathrm{d}^3v=E.
\end{equation}
\end{subequations}
%

We now use the variational method in order to find the extremal of $\mathcal{H}'$, consistent with the restrictions
(\ref{eq:cellrestrictions}) and (\ref{eq:micro}) by using the corresponding Lagrange
multipliers $\{\alpha_M\}$ and $\{\beta_M\}$. This yields
%
\begin{eqnarray}\label{eq:deltaHpdeltafpj}
    \frac{\delta \mathcal{H}'}{\delta f'_J(\vec{v}')} & = & \sum_{M=1}^{K}\int
      \frac{\delta}{\delta f'_J(\vec{v}')}\left[
        f'_M(\vec{v})\ln f'_M(\vec{v})
        \right]
       \mathrm{d}^3v -\sum_{M=1}^{K}\alpha_M\int
       	\frac{\delta f'_M(\vec{v})}{\delta f'_J(\vec{v}')}
      \mathrm{d}^3v\nonumber\\
    & & -\sum_{M=1}^{K}\beta_M\int\epsilon(\vec{v})
    	\frac{\delta f'_M(\vec{v})}{\delta f'_J(\vec{v}')}
      \mathrm{d}^3v \nonumber\\
    & = & \ln f'_J(\vec{v}')+1-\alpha_J-\beta_J \epsilon(\vec{v}')=0.
\end{eqnarray}
%
Solving the last line for $f'_J(\vec{v}')$ renders
%
\begin{comment}
\begin{eqnarray}
    \ln f'_J(\vec{v}') & = & \alpha_J+\beta_J \epsilon(\vec{v}')
       -1\quad\Rightarrow\quad f_J(\vec{v}')\\
    & = & e^{\alpha_J +\beta_J \epsilon(\vec{v}')-1}\nonumber\\
    & = & Ce^{\alpha_J+\beta_J \epsilon(\vec{v}') } \label{eq:relacion1},
\end{eqnarray}
\end{comment}
\begin{equation}\label{eq:relacion1}
	f'_J(\vec{v}') = C\exp\left({\alpha_J+\beta_J \epsilon(\vec{v}') }\right),
\end{equation}
%
where $C$ is a constant. We notice that the use of the variational procedure on
$\mathcal{H}'$ predicts that when the equilibrium is reached, the
distribution function of each cell has the form of the Maxwell-Boltzmann
distribution function, which is consistent with the local equilibrium assumption.

@@

If the complete system is in equilibrium without external forces applied to
the gas, from classical thermodynamics of
systems in equilibrium, we ascertain that the local number
of particles and the local
energy do not depend on the cell number, on a statistical sense, this is
%
\begin{subequations}\label{eq:eqNandE}
\begin{equation}
   \mathcal{N}'_M=\mathcal{N}'\equiv \bar{\mathcal{N}}'
\end{equation}
%
and
%
\begin{equation}
   \quad\mathcal{E}'_M=\mathcal{E}'\equiv\bar{\mathcal{E}}'.
\end{equation}
\end{subequations}
%
In Eqs. (\ref{eq:eqNandE}) the bar implies averaged properties over the complete system.
Moreover, the distribution function is homogeneous, hence $f'_M$
does not depend on the cell number $M$ (\textit{i.e.} $f'_M(\vec v,t)=f'(\vec v,t)$,
$\forall M$). Several properties follow directly from this, \textit{e.g.}
from Eqs.~(\ref{eq:eqNandE}) $E=\sum_M\mathcal{E}'_M=K\bar{\mathcal{E}}'$, $N=\sum_M\mathcal{N}'_M=K\bar{\mathcal{N}}'$. Here we have used Eqs. (\ref{eq:Ecell}) and (\ref{eq:Ncell}).
Eq.~(\ref{eq:CH2}), in terms of Eq.~(\ref{eq:hbfunctional}), can be cast into
%
\begin{equation}
    \mathcal{H}'(t)=\int \sum_{M=1}^{K} [f'(\vec{v},t)\ln f'(\vec{v},t)] \mathrm{d}^{3}v
         = K\int  f'(\vec{v},t) \ln f'(\vec{v},t) \mathrm{d}^{3}v= K H_{B}(t).
\end{equation}
%
In order to identify $\mathcal{H}'$ with the entropy, still we need to show that $\mathcal{H}'(t)$ is extensive,
with respect to $Kf'_(\vec v,t)$.
This is shown by analyzing the following expression
%
\begin{eqnarray}\label{eq:sum-h}
    \int \big[Kf'(\vec{v},t)\big] \ln \big[Kf'(\vec{v},t)\big] \mathrm{d}^{3}v & = & 
        \int \big[(K\ln K)f'(\vec{v},t) + Kf'(\vec{v},t) \ln f'(\vec{v},t)\big]\mathrm{d}^{3}v\nonumber\\
        & = & K\int\Big\{ f'(\vec{v},t)\big[\ln K + \ln f'(\vec{v},t)\big] \Big\}\mathrm{d}^{3}v.
\end{eqnarray}
%
We observe that if the number of particles in the $\mu$-space, $f'(\vec{v},t)$, is much large than
the number of cells $K$, then the first term of Eq.~(\ref{eq:sum-h}) is negligible, and consequently:
%
\begin{equation}\label{aditive-property}
    \int \big[Kf'(\vec{v},t)\big] \ln \big[Kf'(\vec{v},t)\big] \mathrm{d}^{3}v \simeq
    K\int f'(\vec{v},t) \ln f'(\vec{v},t) \mathrm{d}^{3}v,
\end{equation}
%
\textit{i.e.} $\mathcal{H}'_M$ is extensive, and the sum $\sum_M\mathcal{H}'_M$ is the $H$-functional of the complete
system, which reduces to the Boltzmann $H$-functional.
\commb{Hence, $\mathcal{H}'$ can also be identified with a density of entropy for the system.}

Furthermore, in equilibrium the Lagrange multipliers are position- and time-independent,
thus $f'_M(\vec{v})$
reduces to
%
\begin{equation}\label{eq:fbardef}
    f'_M(\vec{v})=C\exp\left(\alpha+\beta \epsilon(\vec{v})\right)
    \equiv \bar{f}'(\vec{v}),\quad M=1,\,\dots\,,K.
\end{equation} 
%
The constant $C$ can be omitted, which is shown by defining the following $\mathcal{H}''$
functional:
%
\begin{equation}
   \mathcal{H}''(t)=\sum_{M=1}^{K}\int \left[f'_M(\vec{v},t)
    \ln f'_M(\vec{v},t)-f'_M(\vec{v},t)\right]\mathrm{d}^3v  \label{CH3},
\end{equation}
%
since $\sum_{M=1}^{K} \int f'_M(\vec{v},t)\mathrm{d}^3v =N$.
$N$ is a constant and we are mainly intersted in the time-derivative of $\mathcal{H}''$, thus
$C$ can be conveniently omitted. In other words, $\mathcal{H}'$ leads the Maxwell-Boltzmann
distribution function for systems in equilibrium.

\subsubsection{Proof of the $H$-theorem for non-homogeneous distributions}

Along this section, we consider a classical gas
whose initial condition is close to the equilibrium,
which ensures that the local equilibrium hypothesis remains valid during
the time-evolution of the system.
Also we will use the following definitions for the deviations of concentration 
and energy from the equilibrium values:
\begin{subequations}\label{eq:restrictionoutsideclassical}
%
\begin{equation}
      \mathcal{N}'_M(t)=\int f'_{M}(\vec{v},t) \mathrm{d}^{3}v=
      \bar{\mathcal{N}}'+\Delta'_M(t)
\end{equation}
%
and
%
\begin{equation}
      \mathcal{E}'_M(t)=\int f'_{M}(\vec{v},t) \epsilon(\vec{v}) \mathrm{d}^{3}v=
      \bar{\mathcal{E}}'+ \delta'_M(t).
\end{equation}
\end{subequations}
%
Here $\bar {\mathcal{N}}' =N/K$ and $\bar{\mathcal{E}}'=E/K$ are
the cell particle number and the cell energy in equilibrium, respectively, they are given by:
%
    \begin{equation}
      \bar{\mathcal{N}}'=
      \int \bar{f}'(\vec{v}) \mathrm{d}^{3}v\quad\textrm{and} \quad
      \bar{\mathcal{E}}'=
      \int \bar{f}'(\vec{v})\epsilon(\vec{v}) \mathrm{d}^{3}v,
    \end{equation}
    %
where we have used $\bar{f}'(\vec{v})$ as defined in Eq.~(\ref{eq:fbardef}).
In Eqs.~(\ref{eq:restrictionoutsideclassical}), $\Delta'_M$ and $\delta'_M$
are considered deviations relative to
$\bar{\mathcal{N}}'$ and $\bar{\mathcal{E}}'$, respectively, and for systems that are
sufficiently close to equilibrium, it is reasonable to expect on the one hand that
$\Delta'_M(t)\ll\bar{\mathcal{N}}'$ and
$\delta'_M(t)\ll\bar{\mathcal{E}}'$, and on the other hand that $\Delta'_M$ and
$\delta'_M$ are sufficiently large compared with fluctuations of $\bar{\mathcal{N}}'$
and $\bar{\mathcal{E}}'$. Under the same perspective, we can assume that every local
distribution function can be written as
\begin{equation}\label{eq:firstorder}
   f'_{M}(\vec{v},t)=\bar{f}'(\vec{v})(1+g'_{M}(\vec{v},t)),\quad
   1\gg|g'_{M}(\vec{v},t)|.
\end{equation}
%
